---
title: "quollr: An R Package for Visalizing 2D Models in High Dimensional Space"
abstract: >
  An abstract of less than 150 words.
draft: true
author:  
  - name: Jayani P.G. Lakshika
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
output: 
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
source("quollr_code.R", local = TRUE)
```

```{r load-libraries}
library(quollr)
```
 
# Introduction

# Methodology

## Usage

dependancies
basic example


### Compute hexagonal bin configurations

### Generate full hex grid

### Construct the 2D model with different options

### Construct the high-D model with different options

### Generate the triangular mesh

### Compute parameter defaults

#### Shift the hexagonal grid origin

#### Benchmark value to remove the low-density hexagons

#### Benchmark value to remove the long edges

### Model function

### Predict 2D embeddings

### Compute residuals

### Visualizations

#### geom_trimesh

#### coloured_long_edges

#### remove long edges

#### show_langevitour

## Tests

# Examples

# Conclusion

# Acknowledgements

This article is created using \CRANpkg{knitr} [@knitr] and \CRANpkg{rmarkdown} [@rmarkdown] in R with the `rjtools::rjournal_article` template. The source code for reproducing this paper can be found at: [https://github.com/JayaniLakshika/paper-quollr](https://github.com/JayaniLakshika/paper-quollr).

