% !TeX root = RJwrapper.tex
\title{quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space}


\author{by Jayani P.G. Lakshika, Dianne Cook, Paul Harrison, Michael Lydeamore, and Thiyanga S. Talagala}

\maketitle

\abstract{%
An abstract of less than 150 words.
}

\begin{verbatim}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)

set.seed(20230531)
\end{verbatim}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{methodology}{%
\section{Methodology}\label{methodology}}

\hypertarget{implementation}{%
\section{Implementation}\label{implementation}}

The package can be installed from CRAN:

\texttt{install.packages("quollr")}

The development version can be installed from GitHub:

\texttt{devtools::install\_github("JayaniLakshika/quollr")}

The following demonstration of the package's functionality assumes \texttt{quollr} has been loaded. We also want to load the built-in data sets \texttt{s\_curve\_noise\_training} and \texttt{s\_curve\_noise\_umap}.

\hypertarget{constructing-the-2d-model}{%
\subsection{Constructing the 2D model}\label{constructing-the-2d-model}}

\hypertarget{preprocessing}{%
\subsubsection{Preprocessing}\label{preprocessing}}

\hypertarget{binning-data}{%
\subsubsection{Binning data}\label{binning-data}}

\hypertarget{obtain-bin-centroids-means}{%
\subsubsection{Obtain bin centroids/ means}\label{obtain-bin-centroids-means}}

\hypertarget{triangulate-bin-centroids-means}{%
\subsubsection{Triangulate bin centroids/ means}\label{triangulate-bin-centroids-means}}

\hypertarget{lifting-the-model-into-high-dimensions}{%
\subsection{Lifting the model into high dimensions}\label{lifting-the-model-into-high-dimensions}}

\hypertarget{tuning-model-parameters}{%
\subsection{Tuning model parameters}\label{tuning-model-parameters}}

\hypertarget{compute-mse}{%
\subsection{Compute MSE}\label{compute-mse}}

\hypertarget{visualizations}{%
\subsection{Visualizations}\label{visualizations}}

\hypertarget{d-model-visualization}{%
\subsubsection{2D model visualization}\label{d-model-visualization}}

\hypertarget{p-d-model-visualization}{%
\subsubsection{p-D model visualization}\label{p-d-model-visualization}}

\hypertarget{prediction}{%
\subsection{Prediction}\label{prediction}}

\hypertarget{tests}{%
\subsection{Tests}\label{tests}}

All functions have tests written and implemented using the \CRANpkg{testthat} (Wickham 2011) in R.

\hypertarget{application}{%
\section{Application}\label{application}}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

This article is created using \CRANpkg{knitr} (Xie 2015) and \CRANpkg{rmarkdown} (Xie, Allaire, and Grolemund 2018) in R with the \texttt{rjtools::rjournal\_article} template. The source code for reproducing this paper can be found at: \url{https://github.com/JayaniLakshika/paper-quollr}.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-testthat}{}}%
Wickham, Hadley. 2011. {``Testthat: Get Started with Testing.''} \emph{The R Journal} 3: 5--10. \url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-knitr}{}}%
Xie, Yihui. 2015. \emph{Dynamic Documents with {R} and Knitr}. 2nd ed. Boca Raton, Florida: Chapman; Hall/CRC. \url{https://yihui.name/knitr/}.

\leavevmode\vadjust pre{\hypertarget{ref-rmarkdown}{}}%
Xie, Yihui, J. J. Allaire, and Garrett Grolemund. 2018. \emph{{R} Markdown: The Definitive Guide}. Boca Raton, Florida: Chapman; Hall/CRC. \url{https://bookdown.org/yihui/rmarkdown}.

\end{CSLReferences}


\address{%
Jayani P.G. Lakshika\\
Monash University\\%
Department of Econometrics and Business Statistics, VIC 3800 Australia\\
%
\url{https://jayanilakshika.netlify.app/}\\%
\textit{ORCiD: \href{https://orcid.org/0000-0002-6265-6481}{0000-0002-6265-6481}}\\%
\email{jayani.piyadigamage@monash.edu}%
}

\address{%
Dianne Cook\\
Monash University\\%
Department of Econometrics and Business Statistics, VIC 3800 Australia\\
%
\url{http://www.dicook.org/}\\%
\textit{ORCiD: \href{https://orcid.org/0000-0002-3813-7155}{0000-0002-3813-7155}}\\%
\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}%
}

\address{%
Paul Harrison\\
Monash University\\%
MGBP, BDInstitute, VIC 3800 Australia\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0002-3980-268X}{0000-0002-3980-268X}}\\%
\href{mailto:paul.harrison@monash.edu}{\nolinkurl{paul.harrison@monash.edu}}%
}

\address{%
Michael Lydeamore\\
Monash University\\%
Department of Econometrics and Business Statistics, VIC 3800 Australia\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0001-6515-827X}{0000-0001-6515-827X}}\\%
\href{mailto:michael.lydeamore@monash.edu}{\nolinkurl{michael.lydeamore@monash.edu}}%
}

\address{%
Thiyanga S. Talagala\\
University of Sri Jayewardenepura\\%
Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka\\
%
\url{https://thiyanga.netlify.app/}\\%
\textit{ORCiD: \href{https://orcid.org/0000-0002-0656-9789}{0000-0002-0656-9789}}\\%
\href{mailto:ttalagala@sjp.ac.lk}{\nolinkurl{ttalagala@sjp.ac.lk}}%
}
