---
title: "quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space"
abstract: >
  An abstract of less than 150 words.
draft: true
author:  
  - name: Jayani P.G. Lakshika
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
output: 
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE)

```

```{r load-libraries}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)

set.seed(20230531)
```

# Introduction

<!--intro with S-curve with 5 methods-->

# Methodology

<!--explain algorithm steps in summary-->

# Implementation

The package can be installed from CRAN:

```install.packages("quollr")```

The development version can be installed from GitHub:

```devtools::install_github("JayaniLakshika/quollr")```

The following demonstration of the package's functionality assumes `quollr` has been loaded. We also want to load the built-in data sets `s_curve_noise_training` and `s_curve_noise_umap`.


<!-- add about main function to gen 2D and highD models-->
## Constructing the 2D model

<!--Discuss the model can be generated with bin centroids or bin means-->

### Preprocessing

<!--Discuss scaling the NLDR data-->

### Binning data 

### Obtain bin centroids/ means

### Triangulate bin centroids/ means

<!-- add about to do hex binning at once-->

## Lifting the model into high dimensions

## Tuning model parameters

## Compute MSE

## Visualizations

### 2D model visualization

<!--2D model vis: triangular mesh-->

### p-D model visualization

<!--highD model vis: langevitour with the model-->

<!--
## Usage

-   dependencies

-   basic example-->

## Prediction

## Tests

All functions have tests written and implemented using the \CRANpkg{testthat} [@testthat] in R.

# Application

<!--select a dataset from cardinalR-->

# Conclusion

# Acknowledgements

This article is created using \CRANpkg{knitr} [@knitr] and \CRANpkg{rmarkdown} [@rmarkdown] in R with the `rjtools::rjournal_article` template. The source code for reproducing this paper can be found at: <https://github.com/JayaniLakshika/paper-quollr>.
